\begin{acknowledgements}
First and foremost, I would like to thank my advisors: Professors Ben Worrell, Jo\"{e}l Ouaknine, and Elias Koutsoupias. I could not have hoped for better supervision during the past three years. This thesis would not exist without their guidance and encouragement.

I am also very grateful to all the wonderful people I have met whilst at Oxford. In particular, I would like to thank Andrew, Clara, Diogo, Edo, Francesco, Francisco, Gon\c{c}alo, Hosein, Hossam, Jasper, Jorge, Luis, Mafalda, Marcelo, Marina, Mathura, Miguel, Nikola, Philippos, Raphael, Rita, Sebouh, Shahab, Smriti, Sophia, Stephen, Straulino, Tom\'{a}s, and Ventsi. Their friendship is invaluable.

Moreover, I would like to thank St Cross College for the diverse and stimulating environment it promotes within its community, which were of great value to me. I will always remember college lunch very fondly.
%I am also thankful to all members of the college football team.

%I am indebted to Professor Gavin Lowe and St Catherine's College for having appointed me as a non-stipendiary lecturer.

%It was an honour to have been examined by Professors Alessandro Abate, Paul Goldberg, Angus MacIntyre, and Jonathan Pila. I thank them for being appreciative of my work and for their encouragement.

Finally, I am obliged to acknowledge that, without the generous funding of the European Research Council (ALGAME grant) and of the Engineering and Physical Sciences Research Council, I would have been unable to conduct the research that led to this thesis.
\end{acknowledgements}
