\section{Turing-degree of MEP}
\label{sec:turing-degree-lics}

Having shown that MEP is undecidable and that MESP reduces to MEP, we are interested in classifying the Turing-degree of MEP.

%TODO: recall Post's theorem in the background section.



\begin{theorem}
\label{thm:turing-degree-2}
  MEP is in $\Sigma_{2}^{0}$.
\end{theorem}

\begin{proof}
  We recall that, by~\cref{thm:Post}, $\emptyset^{(2)}$ (that is, the second Turing jump of the empty set\footnote{Note that any computable set could have been used instead of $\emptyset$.}) is $\Sigma_{2}^{0}$-complete. Below, we present a reduction from MEP to $\emptyset^{(2)}$.

  Consider the function
  \begin{equation*}
    f(\myvector{t}) = \left\| \prod \limits_{i=1}^{k} \exp(A_{i} t_{i}) - C \right\|_{2}^{2}
  \end{equation*}
  and note that $f$ is an exponential-polynomial which has the non-negative reals as co-domain. Clearly, $f$ has a tangential zero if and only if this is a positive instance of MEP.

  For each $n \in \Naturals$, consider the Turing machine $A_{n}$ which does the following: for each $m \in \Naturals$, partition $[0,n]^{k}$ (where $k$ is the number of matrices in the instance of MEP under consideration) in an uniform grid, with mesh size $m^{-1}$, and compute the approximate value of $f$ with error at most $m^{-1}$ at each grid point, and, if it is ever the case that all the approximate values of $f$ are greater than $m^{-1}$, returns false. Note that $A_{n}$ never returns true.

  Now, consider the Turing machine $B$ with access to a Halting Problem oracle (that is, a $\emptyset^{(1)}$ oracle) which, for each $n \in \Naturals$, uses the oracle to decide whether $A_{n}$ terminates and, if that is ever not the case, returns true. Note that $B$ never returns false.

  Finally, note that $B$ terminates if and only if the MEP instance in consideration is positive.
\end{proof}

Moreover, the following result holds.

\begin{theorem}
  If Schanuel's conjecture is true, MEP is $\Sigma_{1}^{0}$-complete.
\end{theorem}

\begin{proof}
  It is clear that MEP is $\Sigma_{1}^{0}$-hard.
  Let $f$ be as in the proof of~\cref{thm:turing-degree-2}. Consider the Turing Machine $T$ which, for each $n \in \Naturals$, uses~\cref{thm:wilkie-macintyre} to decide whether $f$ admits a zero in the region $[0,n]^{k}$. Then $T$ terminates if and only if the instance of MEP under consideration is positive.
\end{proof}
